# Cleaning your data

In this chapter I will show how you can clean your data before you
start any analysis. Doing this before anything else will save you a lot 
of time. It is very frequent that you need to change a variable name or
create a new column in your dataframe. By performing the cleaning and 
modifications before doing any plotting, it helps to ensure you will
be using clean data and it will make your life easier.

## Loading libraries and files

As discussed before, we start by loading the libraries we will use. Later
I will describe what the `stringr` and `janitor` library do. 

```{r, message=FALSE}
library(dplyr)
library(stringr)
library(janitor)
```

And we load up our qPCR data saved from the checkpoint. To do this we
use the function `readRDS`. For this we only specify the path to where
the checkpoint was saved. 

```{r}
qpcr <- readRDS("../checkpoints/loading/qpcr.rds")
```

## Formatting your column names

> Describe the use of janitor here.

## Changing values in a column

> describe the use of mutate from dplyr and the functions from
stringr.

## Checkpoint

>  explain what is being saved and where