---
title: "General analysis"
---


# load data

```{r}
library(tidyverse)
library(janitor)

library(kableExtra)
```

```{r}
qpcr <- read.csv(file = "../data/qPCR.csv")
```


# clean data

```{r}
qpcr <- qpcr %>% janitor::clean_names()
```

# prepare data

```{r}
summarised_qpcr <- qpcr %>% 
    dplyr::group_by(sample_id, group, gene) %>%
    dplyr::summarise(
        mean = mean(ct),
        sd = sd(ct), 
        gene = gene[1], 
        group = group[1],
        sample_id = sample_id[1]
    ) %>% 
    tidyr::separate(
        "group",
        into = c("genotype", "treatment"),
        remove = FALSE
    ) %>%
    dplyr::mutate(
        genotype = factor(genotype, levels = c("WT", "KO")),
        treatment = factor(treatment, levels = c("C", "T"))
    )
```

# plot data

```{r}
summarised_qpcr %>% 
    ggplot2::ggplot(aes(x = mean, y =sd, color = gene, shape = group)) + 
    ggplot2::geom_point() +
    ggplot2::labs(
        x = "Average CT values",
        y = "SD of CT values"
    ) +
    ggplot2::theme_bw()
```

```{r}
summarised_qpcr %>%
    ggplot2::ggplot(aes(y = mean, x = gene, color = group)) + 
    ggplot2::geom_jitter() + 
    ggplot2::labs(x = "Gene", y = "Average CT values") +
    ggplot2::theme_bw()
```

# Normalize data

```{r}
normalized_qpcr <- summarised_qpcr %>%
    tidyr::pivot_wider(
        id_cols = c("sample_id", "group", "treatment", "genotype"),
        values_from = "mean",
        names_from = "gene"
    ) %>% 
    dplyr::mutate(dct = AKT - GAPDH)

normalized_qpcr %>% head
```

# plot normalized data

```{r}
normalized_qpcr %>% 
    ggplot2::ggplot(aes(y = dct, x = group, color = group)) + 
    ggplot2::geom_jitter() + 
    ggplot2::labs(x = "Gene", y = "DCT") +
    ggplot2::theme_bw()
```

Compare ct values with normalized ct values

```{r}
normalized_qpcr %>% 
    ggplot2::ggplot(aes(x = AKT, y = dct, color = group)) +
    ggplot2::geom_smooth(
        method = "lm",
        color = "black", 
        alpha = 0.5, 
        size = .1,
        formula = "y~x"
    ) +
    ggplot2::geom_point() +
    ggplot2::theme_bw()
```

# compare groups

Questions:

1) difference between treatment and control in WT
2) effect of treatment depends on genotype? 

## first question

1) difference between treatment and control in WT

```{r}
analysis_1st <- normalized_qpcr %>% 
    dplyr::filter(grepl("WT", group)) %>%
    dplyr::mutate(group = factor(group, levels = c("WT C", "WT T"))) %>%
    lm(dct ~ group, data = .) %>%
    broom::tidy()
    

analysis_1st %>% 
    kableExtra::kbl() %>%
    kableExtra::kable_classic(full_width = FALSE)
```

To calculate the fold change one needs to flip the sign for the 
estimate, since an increase of CT is actually a decrease in expression.

Fold change is then:

```{r}
2^(-analysis_1st[2, "estimate"])
```

And log fold change is:

```{r}
-analysis_1st[2, "estimate"]
```


```{r}
normalized_qpcr %>% 
    dplyr::filter(grepl("WT", group)) %>%
    dplyr::mutate(group = factor(group, levels = c("WT C", "WT T"))) %>%
    lm(AKT ~ group, data = .) %>%
    broom::tidy() %>%
    kableExtra::kbl() %>%
    kableExtra::kable_classic(full_width = FALSE)
```

## second question

2) effect of treatment depends on genotype? 

```{r}
analysis_2nd <- normalized_qpcr %>% 
    lm(dct ~ genotype*treatment, data = .) %>%
    broom::tidy()
    
analysis_2nd %>% 
    kableExtra::kbl() %>%
    kableExtra::kable_classic(full_width = FALSE)
```

high variability leading to high p-value. 
