# Plotting

In this section we will show how to explore qPCR data through plotting. 
Whenever doing data analysis, it is very important to explore data
through data summaries and plots. This way you learn and get a better
feeling for the generated data. 

If you have been following the previous chapters, the data will
be ready to do the plotting and data exploration. Here
we will introduce the library `ggplot2` and we will combine
with the pivoting functions we previously learned.

```{r, message=FALSE}
library(dplyr)

library(ggplot2)
```

```{r}
qpcr <- readRDS("../checkpoints/cleaning/qpcr.rds")
summarised_qpcr <- readRDS("../checkpoints/preparing/summarised_qpcr.rds")
```

## Visualizing cycle thresholds

The first step is to understand the cycle thresholds obtained in the 
data, even before normalizing them. This is important as we might
find outliers and we check how good the housekeeping genes are.

Moreover, one important step of the analysis is to check the absolute 
values of the cycle thresholds. Sometimes the fold changes
can be misleading if there are high values of cycle thresholds.

Since qPCR data usually is already in the long 
format^[Check the appendix for more details.], we can
directly apply the whole `ggplot2` machinery.

The idea with `ggplot2` is that one can add layers on your plot.
You start with an empty canvas by calling the function
`ggplot` and you specify its aesthetics within the function 
`aes`. Here we want to plot the cycle threshold by sample id, color
by treatment and the shape will be the genotype. 

```{r}
p <- ggplot(
    summarised_qpcr, 
    aes(x = gene, y = mean, color = genotype, shape = treatment)
)
```

```{r}
p + geom_jitter()
```

