# Answering questions

We have been preparing and analysing our data to answer our questions. In
this case we have two: 

1. Is there any difference between treatment and control in the wild type (WT)
condition?
2. Does the effect of treatment depends on genotype? 

The questions will be answered in the next section by using the framework
of linear regression. Linear regression is doing almost the same thing
as a t-test. Also if you want to test normality of the
values this is possible, but since the sample size is very low it is 
very difficult to reject the null hypothesis that the samples do not 
follow a normal distribution. 

The regression framework allows us to extend it to other uses, such as 
using other types of distributions. One example is when we are dealing
with proportions or percentages, usually one would use the 
gamma distribution instead of the normal distribution to model
the data. 

Let's move to answering the questions! 

## First question

Let us recap the question here:

> Is there any difference between treatment and control in the wild type (WT)
condition?

For this we will need to load once again the packages and our previous
checkpoint. R already provides in the base package the function `lm` 
to perform linear regression, so this is actually not necessary to load.

```{r}
library(dplyr)
library(broom)

# we load only the normalized data as this is what will be used when doing
# the comparisons
normalized_qpcr <- readRDS("../checkpoints/normalizing/normalized_qpcr.rds")
```

The code below executes the comparison and prints the output. What we want
to read from the linear regression output is the estimate of the 
slope, that will represent the difference between the two groups! One nice
thing of using linear regression is that you get the difference values between
your conditions, and that might be used to further interpret the results.

```{r}
analysis_first_question <- normalized_qpcr %>% 
    # first filter just for the WT
    dplyr::filter(grepl("WT", group)) %>%
    # Now we factor so the first level is control and the difference
    # then represents treatment - control.
    dplyr::mutate(group = factor(group, levels = c("WT C", "WT T"))) %>%
    # perform the linear regression
    lm(dct ~ group, data = .) %>%
    # tidy up the results so we can display
    broom::tidy()
    

analysis_first_question %>% 
    kableExtra::kbl() %>%
    kableExtra::kable_classic(full_width = FALSE)
```

To calculate the fold change one needs to flip the sign for the 
estimate, since an increase of CT is actually a decrease in expression.
The total fold change is `r 2^(-analysis_first_question[2, "estimate"])` and
log fold change is `r -analysis_first_question[2, "estimate"]`.

The way to calculate the fold change is $2^{(-\text{estimate})}$, where
$\text{estimate}$ corresponds to the second row and second column of the 
table above. And the log fold change is just the negative value of 
$\text{estimate}$. 

The conclusion for this question is that there is indeed a decrease in
AKT expression levels in the treated condition for the WT genotype. We had
already seen this from the plots of the previous chapters.

I would recommend to report this values along with the previous plots,
as the CT values are also an important aspect of the analysis, as an
increase of 30 to 31 is not the same as 16 to 17. 

## Second question

We now move on to the second question.

> Does the effect of treatment depends on genotype? 

Here we are interested in the interaction between the genotype and treatment.
R has an special way to encode this when doing linear regression, by using
the star symbol \*. 

```{r}
analysis_second_question <- normalized_qpcr %>% 
    dplyr::mutate(
        group = factor(group, levels = c("WT C", "WT T", "KO C", "KO T")),
        genotype = factor(genotype, levels = c("WT", "KO"))
    ) %>%
    # notice here the star. we are now interested in the interaction
    lm(dct ~ genotype * treatment, data = .) %>%
    broom::tidy()
    
analysis_second_question %>% 
    kableExtra::kbl() %>%
    kableExtra::kable_classic(full_width = FALSE)
```

We had already seen that actually in the knockout condition there was
a lot of variability. This is reflected in the results here. The last line
of the table above presents the result that interest us, the interaction
between treatment and genotype. We see that the p-value is around 0.71,
so we don't have enough evidence to reject the null hypothesis, which is 
there is no effect in the genotype and treatment. 

# Conclusion

Here we saw how to perform basic statistical analysis of the generated
qPCR data and how to interpret it. The analysis does not stop here,
it is important to also document it. So I will show next how to write a 
report that can be reused for further analysis and whose results
can be distributed to other people.
